---
title: "Chapter 12: Logical Vectors Practice"
name: "Eric Nordmoe"
date: today
format: 
    gfm: 
      toc: true
      toc-depth: 2
      fig-height: 4
      fig-width: 9
editor: visual
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(nycflights13)
```

**1.** Use `mutate()`, `is.na()`, and `count()` together to describe how the missing values in `dep_time`, `sched_dep_time` and `dep_delay` are related

**Answer:**

```{r}

```

**2.** Find all flights where `arr_delay` is missing but `dep_delay` is not. Find all flights where neither `arr_time` nor `sched_arr_time` are missing, but `arr_delay` is.

**Answer:**

```{r}

```


**3.** Examine flight cancellations by time of day and by month. Specifically, determine:  
- At what times of day are cancellations most common? Use the `hour` variable and exclude any flights before 5 am.
- In which months do cancellations occur most frequently?  
- Is there a relationship between the proportion of flights canceled on a given day and the average delay of the flights that did take off?  Use `coord_cartesian()` to limit your plot to include only days where the proportion canceled was less than 0.3.
  
Use `mutate()`, `is.na()`, and `count()` or `summarize()` to explore these patterns. Create visualizations to support your findings.

**Answer:** 

```{r}

```

