---
title: "Chapter 12: Logical Vectors Practice"
name: "Sample Solutions"
format: 
    gfm: 
      toc: true
      toc-depth: 2
      fig-height: 4
      fig-width: 9
editor: visual
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(nycflights13)
```

**1.** Use `mutate()`, `is.na()`, and `count()` together to describe how the missing values in `dep_time`, `sched_dep_time` and `dep_delay` are related

**Answer:**

No scheduled departure times are missing. If a departure time is missing, then its delay is also missing so the flight was probably canceled. 

```{r}
df <- 
  flights |> 
  mutate(
    dep_time_missing = is.na(dep_time),
    dep_delay_missing = is.na(dep_delay),
    sched_dep_time_missing = is.na(sched_dep_time)
  ) 
df |> count(dep_time_missing)
df |> count(dep_delay_missing)
df |> count(sched_dep_time_missing)
```

**2.** Find all flights where `arr_delay` is missing but `dep_delay` is not. Find all flights where neither `arr_time` nor `sched_arr_time` are missing, but `arr_delay` is.

**Answer:**

```{r}
flights  |> 
  filter(is.na(arr_delay) & !is.na(dep_delay))

flights |>
  filter(!is.na(arr_time) & !is.na(sched_arr_time) & is.na(arr_delay))
```

**3.** Examine flight cancellations by time of day and by month. Specifically, determine:  
- At what times of day are cancellations most common? Use the `hour` variable and exclude any flights before 5 am.
- In which months do cancellations occur most frequently?  
- Is there a relationship between the proportion of flights canceled on a given day and the average delay of the flights that did take off?  Use `coord_cartesian()` to limit your plot to include only days where the proportion canceled was less than 0.3.
  
Use `mutate()`, `is.na()`, and `count()` or `summarize()` to explore these patterns. Create visualizations to support your findings.

**Answer:**

* Cancellations are more common later in the day.
* Cancellations most common in February and December.
* A higher proportion cancelled is correlated with longer average delays.

```{r}
# Cancellations by hour of the day
hourly_cancellations <- flights |>
  filter(hour >= 5) |>
  mutate(canceled = is.na(dep_time)) |>
  group_by(hour) |>
  summarize(
    proportion_canceled = mean(canceled)
  ) 

# Plot cancellations by hour
ggplot(hourly_cancellations, aes(x = hour, y = proportion_canceled)) +
  geom_line(color = "blue") +
  labs(title = "Proportion of Canceled Flights by Hour",
       x = "Hour of the Day", y = "Proportion Canceled")

# Cancellations by month
monthly_cancellations <- flights |>
  mutate(canceled = is.na(dep_time)) |>
  group_by(month) |>
  summarize(
    proportion_canceled = mean(canceled)
  )

# Plot cancellations by month
# Following show how to lable months
ggplot(monthly_cancellations, aes(x = factor(month, levels = 1:12, labels = month.abb),
                                  y = proportion_canceled)) +
  geom_col(fill = "red") +
  labs(title = "Proportion of Canceled Flights by Month",
       x = "Month", y = "Proportion Canceled")

# Daily cancellations and delays
daily_cancellations <- flights |>
  mutate(canceled = is.na(dep_time)) |>
  group_by(year, month, day) |>
  summarize(
    proportion_canceled = mean(canceled),
    avg_delay_non_canceled = mean(dep_delay, na.rm = TRUE)
  )

# Scatter plot: Proportion canceled vs. average delay
ggplot(daily_cancellations, aes(x = proportion_canceled, y = avg_delay_non_canceled)) +
  geom_point(alpha = 0.6, color = "purple") +
  coord_cartesian(xlim = c(0, 0.3)) +
  labs(title = "Proportion of Cancellations vs. Average Delay",
       x = "Proportion of Canceled Flights (per day)",
       y = "Average Delay of Non-Canceled Flights (minutes)")
```
