---
title: "Chapter 14: Strings Practice with Solutions"
author: "Sample Solutions"
date: today
format: 
    html: 
      toc: true
      number-sections: true
      number-depth: 3
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

Many data sets have character strings in them. For example, in a file of tweets from Twitter (which are basically just strings of characters), perhaps you want to search for occurrences of a certain word or Twitter handle. Or a character variable in a data set might be location with a city and state abbreviation, and you want to extract those observations with location containing "NY."

In this tutorial, you will learn how to manipulate text data using the package **stringr** and how to match patterns using regular expressions.

# Extracting and locating substrings

```{r}
library(stringr)
fruits <- c("apple", "pineapple", "Pear", "orange", "peach", "banana")
```

**Question 1:** What are the characters in the first through third position of each string in `fruits`?

```{r}
str_sub(fruits, 1, 3)
```

**Question 2:** What happens if you enter a position number greater than the number of characters in the string?

```{r}
str_sub(fruits, 2, 10)
```

R does not complain but returns available characters without error.

**Question 3:** What happens if you enter just a single number? What if it's a negative number?

```{r}
str_sub(fruits, 3)   # From 3rd character to end
str_sub(fruits, -3)  # Last three characters
```

**Question 4:** Is pattern matching case-sensitive?

```{r}
str_detect(fruits, "p")
str_detect(fruits, "P")
```

Yes, it is case-sensitive.

**Question 5:** Which fruits contain "an"?

```{r}
fruits[str_detect(fruits, "an")]
```

**Question 6:** If there is more than one occurrence of the string, which occurrences are returned?

```{r}
str_locate_all(fruits, "an")
```

All occurrences are returned.

**Question 7:** What kind of object is returned by `str_locate_all()`?

```{r}
out <- str_locate_all(fruits, "an")
out
class(out)

```

It returns a list.

**Question 8:** Output the number of these 6 words that contain the string "an".

```{r}
sum(str_detect(fruits, "an"))
```

# Regular expressions

```{r}
str_view(fruits, "^[Pp]")
str_view(fruits, "e$")
```

**Question 9:** Replace all periods with dash characters.

```{r}
info <- c("Home: 507-645-5489", "Cell: 219.917.9871", "My work phone is 507-202-2332", "I don't have a phone")
str_replace_all(info, "\\.", "-")
```

# On your own

```{r}
veggies <- c("carrot", "bean", "peas", "cabbage", "scallion", "asparagus")
str_detect(veggies, "ea")
str_detect(veggies, "s$")
str_detect(veggies, "a.*a")
str_detect(veggies, "^[^c]")
str_locate(veggies, "ca")
```

**Question:** Why does `"^[Ss](.*)(t+)(.+)(t+)"` match "scuttlebutt", "Stetson", and "Scattter", but not "Scatter"?

```{r}
str_view(c("scuttlebutt", "Stetson", "Scattter", "Scatter"), "^[Ss](.*)(t+)(.+)(t+)")
```

The expression requires at least two `t`s separated by one or more characters.

# More Resources
- [Regular expressions with R](http://www.regular-expressions.info/rlanguage.html)
