---
title: "Sample Solutions for Chapter 10:<br>Exploratory Data Analysis Practice"
name: "Sample Solutions"
format: 
    gfm: 
      toc: true
      toc-depth: 2
      fig-height: 4
      fig-width: 9
editor: visual
execute: 
  eval: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(NHANES)
```

**1.** Create a histogram for the `Age` variable from the NHANES (National Health and Nutrition Examination Study ) study. What features do you observe? Is there anything particularly problematic? Try changing the `binwidth` to see if that affects your conclusions. Another option to try is `breaks =` instead of `binwidth`. What do you think `breaks = seq(0, 100, 25)` is doing? Try to use something more sensible than that.

**Answer:**

Perhaps the age variable is truncated at 80. Also, we see quite a few very young children. This group was likely oversampled, perhaps for use in updating growth charts, etc.

```{r}
#try different ways to specify breaks/bars
ggplot(data = NHANESraw, aes(x = Age)) + 
  geom_histogram(binwidth = 5, color = "black", fill = "white")
ggplot(data = NHANESraw, aes(x = Age)) + 
  geom_histogram(breaks = seq(0, 90, 5), color = "black", fill = "white")
# Extreme case, set width to 1
ggplot(data = NHANESraw, aes(x = Age)) + 
  geom_histogram(breaks = seq(0, 90, 1), color = "black", fill = "white")
```

**2.** Modify your previous chart so that height now represents the percent in the corresponding interval. Adjust the binwidth so that you can estimate fairly accurately the percent between the ages of 20 and 30. State your estimate. (Hint: There are several ways to get the percent information. Try to use the `after_stat()`argument with a formula to obtain this.)

**Answer:**

It looks like about 10 percent in the 20-30 age group bar. Note that we did not need to use `mutate()` or summarize to calculate the percents.

```{r}
ggplot(data = NHANESraw, aes(x = Age, y = after_stat(100 * count/sum(count)))) + 
  geom_histogram(breaks = seq(0, 90, 10), color = "black", fill = "white") + 
  labs(y = "Percent") +
  scale_y_continuous(breaks=seq(0,16,1)) # This line is not required but nice
```

**3.** Create separate frequency polygons for the variable `BPSysAve` for each category of the variable `BMI_WHO`. All the frequency polygons should appear on the same graph (not faceted). What are your observations?

**Answer:**

The blood pressures for higher BMI categories tend to be higher. The peaks of the `freq_poly` curves for the higher BMI categories are at higher `BPSysAve` levels. There are far fewer individuals with very low BMI levels, 12.0 to 18.5, than for the other levels. There are quite a few NA values for BMI.

```{r}
ggplot(NHANESraw, aes(x = BPSysAve, color = BMI_WHO)) +
  geom_freqpoly()
```

**4.** Now try repeating the analysis from Question 3 but create a **ridgeline plot** as shown [here](https://r4ds.hadley.nz/layers#sec-geometric-objects:~:text=For%20example%2C%20the%20ggridges,the%20density%20curves%20transparent.) in the R4DS textbook. Do you prefer this view to that in Question 3? Why or why not?

**Answer:**

I prefer this view. It's easier to see that blood pressures tend to be higher for older participants.

```{r}
library(ggridges)
ggplot(NHANES, aes(x = BPSysAve, y = BMI_WHO, fill = BMI_WHO, color = BMI_WHO)) +
  geom_density_ridges(alpha = 0.5, show.legend = FALSE)
```

**5.** In the previous questions, did you notice a substantial number of NA values for the `BMI_WHO` variable? Do some digging and characterize who the participants are who were missing data on this variable. What do most of them have in common? Note: Don't just use `View()` but instead use the data wrangling or graphical tools we have learned.

**Answer:**

Most are very young children and a few very old (over 80). See the frequency table of age below.

```{r}
BMI_missing <- NHANESraw %>%
  filter(is.na(BMI_WHO))

BMI_missing %>% count(Age) %>%
  print(n = Inf)
```
